Recalibrating Gravitational Wave Phenomenological Waveform Model

### 2023-04-21 src/tex/ms.tex
figure fix

\te{I think this discussion needs to be much more precise. Having a waveform model calibrated to particular noise curve could make sense, but its inherently different to just fitting the phase and the amplitude separately like they do in the original paper. So I would recommend that our default is to use a flat PSD and then in addition discuss the one with the PSD. This way, I think we demonstrate more clearly that its the high dimensional fitting that helps, not the just changing the metric used when fitting.}Other than NR waveforms, one need to choose a relevant noise PSD for 
mismatch. We have opted to use a flat PSD for this purpose, as it provides results that are independent of the detector sensitivity and mass scale. The use of a flat PSD ensures that the improvement in accuracy is due solely to the difference in high-dimensional fitting and piece-wise fitting. Furthermore, we are interested in examining the effect of introducing a detector PSD on the optimization process. For this, we have chosen the zero-detuned high-power (\zdethp) noise PSD \citep{aasi2015advanced}. Since the total mass of the system scales with the frequency of the waveform, we must choose a corresponding mass scale to match the frequency range of our noise PSD. We selected a mass scale of $M=50M_{\odot}$ for all waveforms, as this is a commonly observed mass scale in LIGO observations. \te{This choice seems pretty arbitrary. Looking at Fig 8 of 2111.03606, they seem to cluster closer to 60-70. This choice is important and should be more informed.}We will then examine the differences in optimization.

@@

We point out that our treatment to NR waveforms is different from that of \citep{husa2016frequency, khan2016frequency}. In the original calibration process, training waveforms are hybrid waveforms of NR and SpinAlignedEOB
(SEOB) waveforms. The low frequency inspiral part is taken from the SEOB
waveforms while the rest of the waveforms are taken from NR simulations.
Instead, we solely use NR waveforms for comparison, since we are only exploring the possibility of optimizing waveform models. Thus, for simplicity sake, we ignore this procedure. On the other hand, most NR waveforms used (for both training and testing) have long enough time series data, i.e.~$>15$ orbits \citep{boyle2019sxs}, in
which they are long enough to contain part of the inspiral segment and all
merger and ringdown frequency information. We take the frequency limits as
$f_{\mathrm{min}}=0.1f_{\mathrm{RD}}$ and $f_{\mathrm{max}}=1.2f_{\mathrm{RD}}$,	% 0.001 Hz?
where $f_{\mathrm{RD}}$ is the frequency at ringdown. This range covers most of
the IMRPhenomD's frequency range, except the minimum frequency is set higher
than that in the original calibration due to NR length. When compared with
IMRPhenomC, the frequency range is slightly extended to have a higher maximum
frequency. We have the dimensionless frequency spacing $M\Delta f=2.5\times10^{-6}$, 
which is sufficient to capture all features of GW strain.

With the loss function evaluated, we apply gradient descent to optimize the tunable
coefficients as shown in Algorithm~\ref{alg:gradient}. We take $\lambda_i$ to be the 
original coefficients given in \citep{khan2016frequency}. We take them as the initial 
waveform coefficients because they lie in the neighborhood of the minimum that we wish 
to find. Then, by taking $\alpha=10^{-6}$ and $N=30000$, we can optimize the 
coefficients systematically. 
\te{I think we need a bit more discussion about the stopping criterion here. What is N, and why did we choose it to be this number? Ideally we would actually have a plot of the loss function during training. I don't have a good intuition for if its noisy. Does it plateau? If we wanted to ensure that people don't complain about overfitting we could also plot the loss of the validation set to show its not going up.}

@@

We begin by investigating the impact of joint optimization on an individual
waveform. Fig. \ref{fig:0154} illustrates that the optimized waveform
outperforms the original waveform, particularly in the inspiral region, where
the amplitude displays a $50\%$ reduction in error. To ensure fairness, we
selected one of the testing waveforms from the catalog presented in
\citep{khan2016frequency}.
% In general, most waveforms within the neighborhood of the training waveforms
% show improvement in mismatches.

We use a constant noise spectrum, $\mathcal{L}{\mathrm{mean}}$, to evaluate the
mismatch of all testing waveforms, and present the resulting distribution in
Fig. \ref{fig:q148}. The peak of the distribution has shifted towards a lower
mismatch, with a decrease of almost one order of magnitude and a 50\% reduction
in the median. When using $\mathcal{L}{\mathrm{fl}}$, we observe a similar
improvement with a 22.9\% decrease in the median. However, the distribution lacks
a clear peak due to the problem discussed in Section \ref{subsec:optimization}.
Despite this, both distributions exhibit significant improvement, and the
optimization scheme remains unaffected.

% \kw{We then compute the mismatch for all testing waveforms and plot the
% distribution of the mismatch in figure X. We find the distribution...}

\begin{figure}[t]
	\script{q148.py}
	\centering
	\includegraphics[width=\columnwidth]{figures/q148.pdf}
	\caption{Distributions of waveform mismatches calculated using both
	$\mathcal{L}_{\mathrm{mean}}$ and $\mathcal{L}_{\mathrm{fl}}$ in
	recalibration. We use training waveforms listed in Tab.~\ref{tab:q148} and
	mismatches are weighted with the constant noise spectrum. For the
	$\mathcal{L}_{\mathrm{mean}}$ distributions, the median decreased by 50.0\%
	while the median of $\mathcal{L}_{\mathrm{fl}}$ distribution decreased by
	22.9\%.}
	\label{fig:q148}
\end{figure}
\begin{figure}[t]
	\script{q148_q1248_compare.py}
	\centering
	\includegraphics[width=\columnwidth]{figures/q148_q1248_compare.pdf}
	\caption{Distributions of $\log_{10}$ difference in mismatch. The
	distribution labeled \texttt{q148} uses training waveforms listed in
	Tab.~\ref{tab:q148} while the \texttt{q1248} distribution uses waveforms
	listed in Tab.~\ref{tab:q148}~and~\ref{tab:q1248}. Mismatches are calculated
	using the constant noise spectrum with the loss function
	$\mathcal{L}_{\mathrm{mean}}$.}
	\label{fig:q148_q1248_compare}
\end{figure}

Applying the same methods, we observe that the distributions of mismatches
calculated using the {\zdethp} noise spectrum exhibit superior improvement
compared to the weighted mismatch. The shape of the distribution is similar to
that shown in Fig. \ref{fig:q148}. This result is expected, as the IMRPhenomD
model was initially constructed and fitted using the {\zdethp} weighted
mismatch. Consequently, the model is anticipated to closely fit the NR waveforms
while incorporating the influence of the {\zdethp} noise spectrum, rather than a
constant spectrum.

Motivated by the successful improvement of the waveforms, we expand the training
dataset to optimize additional waveforms listed in Tab. \ref{tab:q1248}. The new
set of coefficients generated from this optimization process yields only
marginal improvements in the newly produced waveforms, as seen in Fig.
\ref{fig:q148_q1248_compare}. The high mismatch tail of the optimized
distribution remains comparable in length and endpoint to the original
distribution, indicating that our procedure is incapable of improving these
waveforms. Similarly, utilizing the {\zdethp} spectrum to optimize the loss
function with additional waveforms leads to minimal improvement in the resulting
distribution.

@@

Given that the waveform model's ansatz may not be entirely compatible with NR,
and the optimization procedure is carried out over a distribution of waveforms
with varying source parameters, it is conceivable that some trade-offs in
accuracy exist between different parts of the parameter space. If this is the
cause of the high mismatch tail's failure to improve during joint-optimization,
segmenting the parameter space into smaller subspaces should alleviate this
problem. On the other hand, if the ansatz lacks the correct parameterized form
to capture the NR waveforms' behavior as a function of the intrinsic parameters,
the results will always be biased, and we should not expect any improvement,
even if we segment the parameter space during training.

Since we know intrinsic parameters play an important role in the ansatz, we
would like to investigate how intrinsic parameters affect the recalibration
process. First, we plot the parameter space of $q$ vs. $\chi_{\mathrm{PN}}$ in
Fig. \ref{fig:ps_q148_qchi}. We can see near non-spinning waveforms demonstrate
more consistent improvement, whereas spinning waveforms can either improve or
worsen in terms of mismatch base on their intrinsic parameters. 
%\kw{What does this means?}

Furthermore, we plot the parameter space of $\chi_1$ vs. $\chi_2$ in
Fig.~\ref{fig:ps_q148}. Waveforms along the diagonal axis, i.e.
$\chi_1\approx\chi_2$, show good mismatch improvements. We expect to see this
feature since the original coefficients were fitted using NR waveforms with
equal or similar spin, hence the model prefers waveforms with similar spin. One
interesting feature is how the top-left ($\chi_1<0$ and $\chi_2>0$) and bottom-right ($\chi_1>0$ and $\chi_2<0$) regions respond to
optimization. In the top-left region, waveforms
generally improve with along optimization. However, mismatches in the bottom-right region do not improve after optimization. Most
waveforms even turned worse after optimization. These waveforms correspond to
the waveforms in the high mismatch tail in Fig.~\ref{fig:q148}. \kw{Give justification.}

@@

We divided the parameter space into four regions to analyze the effect of the
recalibration procedure on each region separately (Fig.\ref{fig:ps_q148}). The
training waveforms used for fitting are listed in Table \ref{tab:quadrants}. The
top-left and bottom-right regions have limited data for $q>4$, hence the results are
only valid up to $q\leq4$. From Fig.\ref{fig:ps_q148_quadrant}, we observe
improvements in waveform mismatch for all waveforms, except those in the bottom-right region. For equal-spin waveforms lying in the proximity of the diagonal axis,
the mismatch improves significantly due to most training waveforms being
equal-spin. In the top-left region, the improvement is noticeable, except for a
few defects caused by some testing waveforms having $q>4$. The optimized
waveforms in the bottom-right region have a higher mismatch than the original ones,
indicating poor fitting. Further optimization in a smaller subspace does not
improve the results.

@@

Although our study primarily focused on the IMRPhenomD model, this simple yet
versatile approach can be applied to other differentiable GW models, such as the
IMRPhenomP \citep{hannam2014simple} or IMRPhenomX \citep{pratten2020setting} 
models within the same family. By jointly optimizing a new set of coefficients,
it is expected that both models can be enhanced since they share similar
construction principles to the IMRPhenomD model. For instance, they also use PN
approximants as part of the ansatz in the inspiral segment. Interestingly, while
recalibrating the IMRPhenomX model \citep{pratten2020setting}, which is
parameterized by an additional anti-symmetric spin parameter, it is expected not
to exhibit the degeneracy previously described. A more detailed investigation
may provide valuable insights into the systematics of the Phenom models.
Furthermore, this approach is applicable to other GW model families, such as NR
surrogate models \citep{varma2019surrogate} or EOB models
\citep{taracchini2014effective}. Such an approach could simplify NR waveform
calibration procedures and lead to the improvement of existing models.

% Can we further check systematic bias by injection? But how can we know the
% real mass or spin in the first place if we disregard GW models.

@@

In this paper, we have presented a systematic method to recalibrate GW models.
This method utilizes {\jax}'s automatic differentiation to apply
derivative-based optimization to recalibrate GW models jointly. Using the new
implementation of the IMRPhenomD model, {\ripple}, which is written in \jax, in
conjunction with NR waveforms from the SXS catalog, we recalibrate waveform
coefficients of the IMRPhenomD model. In general, the waveform accuracy can be
improved by 50\%. Comparing {\zdethp} weighted and unweighted mismatch, weighted
mismatches have a slightly better improvement. In contrast, different types of
loss function result in significantly different final mismatch distributions,
where the result can be seen in Fig.~\ref{fig:q148}. By increasing the number of
training waveforms, we see a slight improvement increase in
Fig.~\ref{fig:q148_q1248_compare}.

Furthermore, we investigated how the intrinsic parameters affect the
improvement. Fig.~\ref{fig:ps_q148} shows that the optimization procedure has a
certain preference for waveforms lying in the top-left region while the fourth
quadrant cannot be improved. To further test this result, we recalibrate
waveforms in separate regions in parameter space. As shown in
Fig.~\ref{fig:ps_q148_quadrant}, this recalibration process gives further
improvement to the top-left region while the bottom-right region shows similar
result. This indicates that the model's ansatz does not fit waveforms in this region. This phenomenon is due to the reduced spin approximation used
in parameterizing the ansatz, where degeneracies between $\chi_1$ and $\chi_2$
are introduced.

>>>>

%\te{I think this discussion needs to be much more precise. Having a waveform model calibrated to particular noise curve could make sense, but its inherently different to just fitting the phase and the amplitude separately like they do in the original paper. So I would recommend that our default is to use a flat PSD and then in addition discuss the one with the PSD. This way, I think we demonstrate more clearly that its the high dimensional fitting that helps, not the just changing the metric used when fitting.}
Other than NR waveforms, one need to choose a relevant noise PSD for 
mismatch. We have opted to use a flat PSD for this purpose, as it provides results that are independent of the detector sensitivity and mass scale. The use of a flat PSD ensures that the improvement in accuracy is due mainly to the difference in high-dimensional fitting and piece-wise fitting, but not due to the use of a different mass scale. Furthermore, we are interested in examining the effect of introducing a detector PSD on the optimization process. For this, we have chosen the zero-detuned high-power (\zdethp) noise PSD \citep{aasi2015advanced}. Since the total mass of the system scales with the frequency of the waveform, we must choose a corresponding mass scale to match the frequency range of our noise PSD. We selected an arbitrary mass scale of $M=50M_{\odot}$ for all waveforms for demonstration, as this is a commonly observed mass scale in LIGO observations. 
%\te{This choice seems pretty arbitrary. Looking at Fig 8 of 2111.03606, they seem to cluster closer to 60-70. This choice is important and should be more informed.}We will then examine the differences in optimization.

@@

We point out that our treatment to NR waveforms is different from that of \citep{husa2016frequency, khan2016frequency}. In the original calibration process, training waveforms are hybrid waveforms of NR and SpinAlignedEOB
(SEOB) waveforms. The low frequency inspiral part is taken from the SEOB
waveforms while the rest of the waveforms are taken from NR simulations.
Instead, we solely use NR waveforms for comparison, since we are only exploring the possibility of optimizing waveform models. Thus, for simplicity sake, we ignore this procedure. On the other hand, most NR waveforms used (for both training and testing) have long enough time series data, i.e.~$>15$ orbits \citep{boyle2019sxs}, in
which they are long enough to contain part of the inspiral segment and all
merger and ringdown frequency information. We take the frequency limits as
$f_{\mathrm{min}}=0.1f_{\mathrm{RD}}$ and $f_{\mathrm{max}}=1.2f_{\mathrm{RD}}$,	
where $f_{\mathrm{RD}}$ is the frequency at ringdown. This range covers most of
the IMRPhenomD's frequency range, except the minimum frequency is set higher
than that in the original calibration due to NR length. When compared with
IMRPhenomC, the frequency range is slightly extended to have a higher maximum
frequency. We have the dimensionless frequency spacing $M\Delta f=2.5\times10^{-6}$, 
which is sufficient to capture all features of GW strain.

With the loss function evaluated, we apply gradient descent to optimize the tunable
coefficients as shown in Algorithm~\ref{alg:gradient}. We take $\lambda_i$ to be the 
original coefficients given in \citep{khan2016frequency}. We take them as the initial 
waveform coefficients because they lie in the neighborhood of the minimum that we wish 
to find. Then, by taking $\alpha=10^{-6}$, we see in Fig. \ref{fig:loss} plateau at 
around $N=30000$, hence we end the optimization here.  
%\te{I think we need a bit more discussion about the stopping criterion here. What is N, and why did we choose it to be this number? Ideally we would actually have a plot of the loss function during training. I don't have a good intuition for if its noisy. Does it plateau? If we wanted to ensure that people don't complain about overfitting we could also plot the loss of the validation set to show its not going up.}

@@

We begin by investigating the impact of joint optimization on an individual
waveform. Fig. \ref{fig:0154} illustrates that the optimized waveform
outperforms the original waveform, particularly in the inspiral region, where
the amplitude displays a $50\%$ reduction in error. To ensure fairness, we
selected one of the testing waveforms from the catalog presented in
\citep{khan2016frequency}.

We use the constant PSD weighted loss function, $\mathcal{L}_{\mathrm{mean}}$, to evaluate the
mismatch of all testing waveforms, and present the resulting distribution in
Fig. \ref{fig:q148}. The peak of the distribution has shifted towards a lower
mismatch, with a decrease of almost one order of magnitude and a 50\% reduction
in the median. When using $\mathcal{L}_{\mathrm{fl}}$, we observe a similar
improvement with a 22.9\% decrease in the median. However, the distribution lacks
a clear peak due to the problem discussed in Section \ref{subsec:optimization}.
Despite this, both distributions exhibit significant improvement, and the
optimization scheme remains unaffected.

\begin{figure}[t]
	\script{q148.py}
	\centering
	\includegraphics[width=\columnwidth]{figures/q148.pdf}
	\caption{Distributions of waveform mismatches calculated using both
	$\mathcal{L}_{\mathrm{mean}}$ and $\mathcal{L}_{\mathrm{fl}}$. We use 
	training waveforms listed in Tab.~\ref{tab:q148} and
	mismatches are weighted with the constant PSD. For the
	$\mathcal{L}_{\mathrm{mean}}$ distributions, the median decreased by 50.0\%
	while the median of $\mathcal{L}_{\mathrm{fl}}$ distribution decreased by
	22.9\%.}
	\label{fig:q148}
\end{figure}
\begin{figure}[t]
	\script{q148_q1248_compare.py}
	\centering
	\includegraphics[width=\columnwidth]{figures/q148_q1248_compare.pdf}
	\caption{Distributions of $\log_{10}$ difference in mismatch. The
	distribution labeled \texttt{q148} uses training waveforms listed in
	Tab.~\ref{tab:q148} while the \texttt{q1248} distribution uses waveforms
	listed in Tab.~\ref{tab:q148}~and~\ref{tab:q1248}. Mismatches are calculated
	using the constant PSD with the loss function $\mathcal{L}_{\mathrm{mean}}$.}
	\label{fig:q148_q1248_compare}
\end{figure}

Applying the same methods, we observe that the distributions of mismatches
calculated using the {\zdethp} PSD exhibit superior improvement
compared to the weighted mismatch. The shape of the distribution is similar to
that shown in Fig. \ref{fig:q148}. This result is expected, as the IMRPhenomD
model was initially constructed and fitted using the {\zdethp} weighted
mismatch. Consequently, the model is anticipated to closely fit the NR waveforms
while incorporating the influence of the {\zdethp} PSD, rather than a
constant PSD.

Motivated by the successful improvement of the waveforms, we expand the training
dataset to optimize additional waveforms listed in Tab. \ref{tab:q1248}. The new
set of coefficients generated from this optimization process yields only
marginal improvements in the newly produced waveforms, as seen in Fig.
\ref{fig:q148_q1248_compare}. The high mismatch tail of the optimized
distribution remains comparable in length and endpoint to the original
distribution, indicating that our procedure is incapable of improving these
waveforms. Similarly, utilizing the {\zdethp} PSD to optimize the loss
function with additional waveforms leads to minimal improvement in the resulting
distribution.

@@

Given that the waveform model's ansatz may not be entirely compatible with NR,
and the optimization procedure is carried out over a distribution of waveforms
with varying intrinsic parameters, it is conceivable that some trade-offs in
accuracy exist between different parts of the parameter space. If this is the
cause of the high mismatch tail's failure to improve during joint-optimization,
segmenting the parameter space into smaller subspaces should alleviate this
problem. On the other hand, if the ansatz lacks the correct parameterized form
to capture the NR waveforms' behavior as a function of the intrinsic parameters,
the results will always be biased, and we should not expect any improvement,
even if we segment the parameter space during training.

Since we know intrinsic parameters play an important role in the ansatz, we
would like to investigate how intrinsic parameters affect the recalibration
process. First, we plot the parameter space of $q$ vs. $\chi_{\mathrm{PN}}$ in
Fig. \ref{fig:ps_q148_qchi}. We can see near non-spinning waveforms demonstrate
more consistent improvement, since the ansatz are developed base on non-spinning 
waveforms. Also, the original coefficients were fitted using NR waveforms with
equal or similar spin, hence the model prefers waveforms with similar spin. On 
the other hand, spinning waveforms can either improve or worsen in terms of 
mismatch base on their intrinsic parameters. To further discuss the behavior of 
non-spinning waveforms, we plot the parameter space of $\chi_1$ vs. $\chi_2$ in
Fig.~\ref{fig:ps_q148}. Waveforms along the diagonal axis, i.e. 
$\chi_1\approx\chi_2$, show good mismatch improvements as discussed above. 
Meanwhile, the top-left ($\chi_1<0$ and $\chi_2>0$) and bottom-right ($\chi_1>0$ 
and $\chi_2<0$) regions respond to optimization differently. In the top-left region, waveforms
generally improve with along optimization. However, mismatches in the bottom-right 
region do not improve after optimization. Most waveforms even turned worse after 
optimization.

@@

We divided the parameter space into four regions to analyze the effect of the
recalibration procedure on each region separately (Fig.~\ref{fig:ps_q148}). The
training waveforms used for fitting are listed in Table~\ref{tab:quadrants}. The
top-left and bottom-right regions have limited data for $q>4$, hence the results are
only valid up to $q\leq4$. From Fig.~\ref{fig:ps_q148_quadrant}, we observe
improvements in waveform mismatch for all waveforms, except those in the bottom-right 
region. For equal-spin waveforms lying in the proximity of the diagonal axis,
the mismatch has additional improvement due to most training waveforms being
equal-spin. In the top-left region, the improvement is significant, except for a
few defects caused by some testing waveforms having $q>4$. This can be seen clearer in 
Fig.~\ref{fig:all_quadrants}, which we see that the optimized histogram shifts downward uniformly. 
The optimized waveforms in the bottom-right region have a higher mismatch than the original ones,
indicating poor fitting. These waveforms correspond to the waveforms in the high mismatch tail 
in Fig.~\ref{fig:q148}, which we can see in Fig.~\ref{fig:all_quadrants}. Thus, we show that 
further optimization in a smaller subspace does not improve the results.

@@

Although our study primarily focused on the IMRPhenomD model, this simple yet
versatile approach can be applied to other differentiable GW models, such as the
IMRPhenomP \citep{hannam2014simple} or IMRPhenomX \citep{pratten2020setting} 
models within the same family. By jointly optimizing a new set of coefficients,
it is expected that both models can be enhanced since they share similar
construction principles to the IMRPhenomD model. For instance, they also use PN
approximants as part of the ansatz in the inspiral segment. It will be interesting 
to recalibratethe IMRPhenomX model \citep{pratten2020setting}. Because it is
parameterized by an additional anti-symmetric spin parameter, it is expected not
to exhibit the degeneracy previously described. A more detailed investigation
may provide valuable insights into the systematics of the Phenom models.
Furthermore, this approach is applicable to other GW model families, such as NR
surrogate models or EOB models introduced previously. Such an approach could 
simplify NR waveform calibration procedures and lead to the improvement of 
existing models.

@@

In this paper, we have presented a systematic method to recalibrate GW models.
This method utilizes {\jax}'s automatic differentiation to apply
derivative-based optimization to recalibrate GW models jointly. Using the new
implementation of the IMRPhenomD model, {\ripple}, which is written in \jax, in
conjunction with NR waveforms from the SXS catalog, we recalibrate waveform
coefficients of the IMRPhenomD model. In general, the waveform accuracy can be
improved by 50\%. Comparing {\zdethp} weighted and unweighted mismatch, weighted
mismatches have a slightly better improvement. In contrast, different types of
loss function result in significantly different final mismatch distributions. As 
seen in Fig.~\ref{fig:q148}, $\mathcal{L}_{\mathrm{mean}}$ outperforms 
$\mathcal{L}_{fl}$. By increasing the number of training waveforms, we see a 
slight improvement increase in Fig.~\ref{fig:q148_q1248_compare}.

Furthermore, we investigated how the intrinsic parameters affect the
improvement. Fig.~\ref{fig:ps_q148} shows that the optimization procedure has a
certain preference for waveforms lying in the top-left region while the bottom-right 
region cannot be improved. To further test this result, we recalibrate
waveforms in separate regions in parameter space. From Fig.~\ref{fig:ps_q148_quadrant}, 
we can see that this recalibration process gives further
improvement to the top-left region while the bottom-right region shows similar
result. This indicates that the ansatz does not fit waveforms in this region. 
This phenomenon is due to the reduced spin approximation used in parameterizing the 
ansatz, where degeneracies between $\chi_1$ and $\chi_2$ are introduced.
